Interpretable Neural Networks using EAGGA

### 2025-03-13 paper/main.tex
EAGGA bullets

\subsection{Hyperparameter Optimization (HPO)}
- in contrast to model parameters, which are optimised during training, hyperparameters are parameters describing the ML algo + are specified before training
- still often have great impact on final (trained) model performance -> can be optimised too
- formal definition
  * dataset $\mathcal{D}\subseteq\mathcal{X}\times\mathcal{Y}$ consisting of $n$ tuples $(\boldsymbol{x}^{(i)}, y^{(i)})\stackrel{i.i.d.}{\sim}\mathbb{P}_{\boldsymbol{x}y}$
    (data-generating distribution), i.e. $\mathcal{D}=\{(\boldsymbol{x}^{(i)}, y^{(i)})\}_{i=1}^n$
  * ML algo $\mathcal{I}(\cdot,\boldsymbol\lambda)$ with HP config $\boldsymbol\lambda\in\Lambda$ maps dataset to fitted model
    $f:\mathcal{X}\rightarrow\mathbb{R}^g\subseteq\mathcal{Y}$ (where $g\in\mathbb{N}$ is \# of classes of target),
    i.e. $\mathcal{I}:(\mathbb{D}\times\Lambda)\rightarrow\mathcal{H}$, with $\mathbb{D}$ space of all finite datasets and $\mathcal{H}$ hypothesis space (space of all models $f$)
  * denote $\mathcal{I}_{\boldsymbol\lambda}$ as (untrained) model with fixed HP config $\boldsymbol\lambda$
  * loss function $L$
  * goal of model training: for fixed $\boldsymbol\lambda$, have $\mathcal{I}$ find model $f$ minimising expected generalisation error
    $GE(\mathcal{I},\boldsymbol\lambda,\mathcal{D},L)=\mathbb{E}_{(\boldsymbol{x},y)\sim\mathbb{P}_{\boldsymbol{x}y}}[L(y,\mathcal{I}_{\boldsymbol\lambda}(\mathcal{D})(\boldsymbol{x}))]$
  * goal of HP optim: find $\boldsymbol\lambda$ minimising expected generalisation error, i.e. $\argmin_{\boldsymbol\lambda\in\Lambda} GE(\mathcal{I},\boldsymbol\lambda,\mathcal{D},L)$
- \citep[p. 3]{10.1145/3610536}
- HPO = black-box optim problem -> we can apply any black box algo to tune

@@

  * vector of objectives $c_1,c_2,...,c_m=\boldsymbol{c}:\Lambda\rightarrow\mathbb{R}^m$

@@

- Pareto set: set of nondominated points $\mathcal{P}:=\{\boldsymbol\lambda\in\Lambda|\nexists\boldsymbol\lambda'\in\Lambda\text{ s.t. }\boldsymbol\lambda'\prec\boldsymbol\lambda\}$

@@

    - $\argmin_{\boldsymbol\lambda\in\Lambda} w_i c_i(\boldsymbol\lambda)$ with $\sum_{i=1}^k w_i=1$ and $w_i>0,\forall i=1,...,k$

@@

    - $\text{w.l.o.g.} \argmin_{\boldsymbol\lambda\in\Lambda} c_1(\boldsymbol\lambda) \text{ s.t. } c_2(\boldsymbol\lambda)\le\epsilon_2,...,c_m(\boldsymbol\lambda)\le\epsilon_m$

@@

%likely shouldn't be an entire section
\section{EAGGA}
- combines all previously mentioned topics: interpretability, HPO, MOO
- for our problem (interpretability) we cannot use regular EA algorithms as we are also interested in reducing pairwise interactions + non-monotone feature effects
- this creates high-dim search space as shown in EAGGA paper -> they provide a framework to efficiently traverse through a space that is reduced to only sensible HP
  configuration (e.g. interaction of A,B with A monot incr and B monot decr -> couldn't be guaranteed -> don't look at it in first place)
- NSGA-II inspired evolutionary / genetic algorithm
- introduces group structure for more efficient optimization over the entire search space

@@

>>>>

\subsection{Hyperparameter Optimization (HPO)}
- in contrast to model parameters, which are optimised during training, hyperparameters are parameters describing the ML algo + are specified before training
- still often have great impact on final (trained) model performance -> can be optimised too
- formal definition
  * dataset $\mathcal{D}\subseteq\mathcal{X}\times\mathcal{Y}$ consisting of $n$ tuples $(\boldsymbol{x}^{(i)}, y^{(i)})\stackrel{i.i.d.}{\sim}\mathbb{P}_{\boldsymbol{x}y}$
    (data-generating distribution), i.e. $\mathcal{D}=\{(\boldsymbol{x}^{(i)}, y^{(i)})\}_{i=1}^n$
  * ML algo $\mathcal{I}(\cdot,\boldsymbol\lambda)$ with HP config $\boldsymbol\lambda\in\boldsymbol\Lambda$ maps dataset to model
    $\hat{f}:\mathcal{X}\rightarrow\mathbb{R}^g\subseteq\mathcal{Y}$ (where $g\in\mathbb{N}$ is \# of classes of target),
    i.e. $\mathcal{I}:(\mathbb{D}\times\boldsymbol\Lambda)\rightarrow\mathcal{H}$, with $\mathbb{D}$ space of all finite datasets and $\mathcal{H}$ hypothesis space (space of all models $\hat{f}$)
  * denote $\mathcal{I}_{\boldsymbol\lambda}$ as (untrained) model with fixed HP config $\boldsymbol\lambda$
  * loss function $L$
  * goal of model training: for fixed $\boldsymbol\lambda$, have $\mathcal{I}$ find model $\hat{f}$ minimising expected generalisation error
    $GE(\mathcal{I},\boldsymbol\lambda,\mathcal{D},L)=\mathbb{E}_{(\boldsymbol{x},y)\sim\mathbb{P}_{\boldsymbol{x}y}}[L(y,\mathcal{I}_{\boldsymbol\lambda}(\mathcal{D})(\boldsymbol{x}))]$
  * goal of HP optim: find $\boldsymbol\lambda$ minimising expected generalisation error, i.e. $\argmin_{\boldsymbol\lambda\in\boldsymbol\Lambda} GE(\mathcal{I},\boldsymbol\lambda,\mathcal{D},L)$
- \citep[p. 3]{10.1145/3610536}
- HPO = black-box optim problem -> we can apply any black box algo to tune

@@

  * vector of objectives $c_1,c_2,...,c_m=\boldsymbol{c}:\boldsymbol\Lambda\rightarrow\mathbb{R}^m$

@@

- Pareto set: set of nondominated points $\mathcal{P}:=\{\boldsymbol\lambda\in\boldsymbol\Lambda|\nexists\boldsymbol\lambda'\in\boldsymbol\Lambda\text{ s.t. }\boldsymbol\lambda'\prec\boldsymbol\lambda\}$

@@

    - $\argmin_{\boldsymbol\lambda\in\boldsymbol\Lambda} w_i c_i(\boldsymbol\lambda)$ with $\sum_{i=1}^k w_i=1$ and $w_i>0,\forall i=1,...,k$

@@

    - $\text{w.l.o.g.} \argmin_{\boldsymbol\lambda\in\boldsymbol\Lambda} c_1(\boldsymbol\lambda) \text{ s.t. } c_2(\boldsymbol\lambda)\le\epsilon_2,...,c_m(\boldsymbol\lambda)\le\epsilon_m$

@@

%likely shouldn't be an entire section
\section{EAGGA}
- NSGA-II inspired evolutionary + genetic algorithm
- uses AUC-ROC as performance metric and 3 interpretation metrics
  * NF: rel. \# of features used in model
  * NI: rel. \# of pairwise feature interaction effects
  * NNM: rel. \# of non-monotone feature effects
- aims to find high performing models (AUC) with low NF, low NI, and NNM -> ensures better interpretability than simply optimising for performance
  but creates very large objective space $\check{\boldsymbol\Lambda}$ to be optimised over with
  tuples $(\boldsymbol\lambda, \boldsymbol{s}, \boldsymbol{I}_s, \boldsymbol{m}_{\boldsymbol{I}_{\boldsymbol{s}}})=\check{\boldsymbol\lambda}\in\check{\boldsymbol\Lambda}$
  with
  * $\boldsymbol\lambda\in\boldsymbol\Lambda$ ``regular'' model HP config
  * $\boldsymbol{s}\in\{0,1\}^p$ vector denoting feature usage
  * $\boldsymbol{I}_s\in\{0,1\}^{p\times p}$ matrix denoting pairwise interactions
  * $\boldsymbol{m}_{\boldsymbol{I}_{\boldsymbol{s}}}\in\{-1,0,1\}^p$ vector denoting monotonicity constraint of each feature (-1 decreasing, 0 none, 1 increasing)
- thus would require to solve $\argmin_{\check{\boldsymbol\lambda}\in\check{\boldsymbol\Lambda}} (GE(\mathcal{I}_{\check{\boldsymbol\lambda}},\mathcal{D}),NF(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}}),NI(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}}),NNM(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}}))$
- \citep[pp. 540f]{EAGGA}
- introduces equivalence relation $R$ ``allowed to interact'' on features that significantly reduces search space for more efficient optimization
- thus: augmented search space $\tilde{\boldsymbol\Lambda}=\boldsymbol\Lambda\times\mathcal{G}$ comprising model HP space and group structure space
  * each group structure $\boldsymbol{G}\in\mathcal{G}$ consists of $g$ groups
    - $G_1$ set of excluded features
    - $\forall k=2,...,g:G_k=(E_k,M_{E_k})$ tuple
      * $E_k$ set of features allowed to interact with each other
      * $M_{E_k}$ monotonicity constraint of entire group $k$
- EA on model hyperparameters $\boldsymbol\Lambda$
- GGA on group structure space $\mathcal{G}$
- \citep[pp. 541f]{EAGGA}
- futher introduces special initialisation of group structures to increase sample-efficiency of EAGGA algorithm compared to random init
- \citep[pp. 542f]{EAGGA}

@@

- original EAGGA applied to XGBoost model proved to vastly outperform union of competitor models w.r.t. dominated hypervolume and comparable or better than ParEGO on
  extended search space $\check{\boldsymbol\Lambda}$ \citep[pp. 543-545]{EAGGA}
